---
title: "Modeling PHQ-9 Scores in Chilean Longitudinal Social Survey"
author: "Carlos Delgado G."
date: "13-11-2021"
output:
  ioslides_presentation:
    css: css_presentacion.css


---



```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(pscl)
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning = FALSE)
```

```{r}
#Recodificaciones
source('Recodificaciones.R', encoding = 'UTF-8', echo=FALSE)
source('funciones midap.R', encoding = 'UTF-8', echo=FALSE)
```

## About my research

- **Broad Interest**: Asses which groups of people have the more severe and persistent depressive symptoms (as measured by PHQ-9).
- **Research goal**: To model the longitudinal variations of PHQ-9 scores among ELSOC participants.
- **Metodology**: Generalized linear mixed effects models.
- **Current state**: Exploring probability distributions for the first year of the study.
- **Today's Presentation**: A discrete data probability model that take into account the excess of zeros (ZINB regression).

## About the data

- [ELSOC-COES](https://coes.cl/encuesta-panel/) Chilean Longitudinal Social Survey. 5 year panel data of adults since 2016. Aims to be representative of urban chilean population.
- More than 300 variables from different modules: *"Territory"*, *"Social attitudes and networks"*, *"Citizenship and Democracy"*,*"Inequality and Legitimacy"*, *"Social Conflict"*, *"Health and Wellbeing"*, *"Sociodemographic"*.
- **Health and Wellbeing**: Life satisfaction, PHQ-9, Stressors, recient health issues, substances use, support networks, acces to health systems, BMI.
- You have PHQ-9 scores for the same group of adults each wave of the study, with hundreds of variables to play with. 

# PHQ-9 Scores in ELSOC respondents (First wave)

## PHQ-9 Items (1-3)

Over the last 2 weeks, how often have you been bothered
by any of the following problems? 

```{r}
grafo_phq(c("s11_01_rec","s11_02_rec","s11_03_rec"),c("Little interest in things","Feeling down","Trouble with sleep"))
```

## PHQ-9 Items (4-6)

Over the last 2 weeks, how often have you been bothered
by any of the following problems? 

```{r}
grafo_phq(c("s11_04_rec","s11_05_rec","s11_06_rec"),c("Tired or little energy","Poor appetite or overeating","Feeling bad about yourself"))
```

## PHQ-9 Items (7-9)

Over the last 2 weeks, how often have you been bothered
by any of the following problems? 

```{r}
grafo_phq(c("s11_07_rec","s11_08_rec","s11_09_rec"),c("Trouble concentrating","Moving slowly or restless","Thoughts of hurting yourself"))
```


## Aggregated PHQ-9 score {.smaller}

For each of the 9 items recode "Not at all" as 0, "Several days" as 1, "More than half the days" as 2, and "Nearly everyday" as 3. Then sum all items. The result is a score that ranges from 0 to 27.

```{r}
ggplot(elsoc_2016,aes(x=s11_phq9))+
  geom_bar(fill="cornflowerblue",color="white")+
  labs(x="PHQ-9 Score",
       y="Frecuency")+
  theme_classic()
```

## Differences in depressive symptoms (1)

```{r}
phq_boxplot("m0_sexo_fac","PHQ-9 Scores by gender",etiq = c("Male", "Female"))
```

## Differences in depressive symptoms (2)

```{r}
phq_boxplot("m02_fac", "PHQ-9 Scores by Occupation", etiq = c("Employed","Domestic Labour","Retired","Unemployed","Other Categories"))
```

## Differences in depressive symptoms (3)
```{r}
phq_boxplot("s_imc_fac","PHQ-9 Scores by BMI categories",etiq = c("Normal","Overweight","Obese"))
```

## Differences in depressive symptoms (4)
```{r}
phq_boxplot("s08_fac","PHQ-9 scores between Smokers and Non Smokers",etiq = c("Non smoker","Smoker"))
```

## Differences in depressive symptoms (5)
```{r}
phq_boxplot("s01_fac","PHQ-9 scores by life satisfaction",etiq = c("Unsatisfied","Neither satisfied nor unsatisfied","Satisfied"))
```

# How can we model discrete over-disperse data?

## Modeling approaches for discrete over-disperse data {.smaller}

<div style="float: left; width: 30%;">
- Against the normal model
- The Poisson Model
- Over disperse data (Negative Binomial)
- Excess of ceros (Zero-Inflated)
</div>

<div style="float: right; width: 70%;">

```{r,fig.width=6}
ggplot(elsoc_2016,aes(x=s11_phq9))+
  geom_bar(fill="cornflowerblue",color="white")+
  labs(x="PHQ-9 Score",
       y="Frecuency")+
  theme_classic()
```

</div>

# Zero-Inflated Negative Binomial Regression

##  ZINB Regression

- You model two separate processes:
-     PHQ-9 score > 1
-     Having a 0 PHQ-9 score

## Toy example ZINB Regression {.smaller}

- **Gender**: Categorical: *"Male"*, *"Female"*.
- **Stressors**: Number of vital events in the last year (death of family member, illnes, Job layoff, etc). Numeric 0-8.
- **Age**: Numeric 18-88.
- **Smoker**: Categorical: *"Non Smoker"*, *"Smoker"*.
- **Body Mass Index (BMI)**: Continuous, median value 26,83.
- **Life Satisfaction**: Categorical: *"Not Satisfied"* (Negative),*"Neither satisfied nor dissatisfied"* (Neutral), *"Satisfied"* (Positive).

## Modeling PHQ-9 Scores by ZINB regression {.reduced}


```{r}
mod_zinb<- zeroinfl(s11_phq9~s_imc+m0_sexo_fac+s13_estresores+m0_edad+s01_fac+s08_fac,data = elsoc_2016,dist="negbin")

```

```{r,results='asis',fig.height=5}
texreg::htmlreg(mod_zinb,single.row=TRUE,
                custom.coef.names=c("Count: Intercept","Count: BMI","Count: Female","Count: Stressors",
                                                             "Count: Age","Count: Neutral Satisfaction","Count: Positive Satisfaction",
                                                             "Count: Smoker","Count: Over dispersion parameter",
                                                             "Zero: Intercept","Zero: BMI","Zero: Female","Zero: Stressors",
                                                             "Zero: Age","Zero: Neutral Satisfaction","Zero: Positive Satisfaction",
                                                             "Zero: Smoker"),
                custom.model.name="Zero-Inflated Negative Binomial Regression")
```

## Coefficients Interpretation {.smaller}

```{r}
exp_coef<- exp(coefficients(mod_zinb))
tabla_coef <- tibble("Variable"=unique(gsub(x=names(exp_coef),pattern = "count_|zero_",replace="")),
       "Count"=matrix(exp(coefficients(mod_zinb)),ncol = 2)[,1],
       "Zero"=matrix(exp(coefficients(mod_zinb)),ncol = 2)[,2])
tabla_coef$Variable <- c("Intercept","BMI","Female","Stressors","Age","Neutral Satisfaction","Positive Satisfaction","Smoker")
  
knitr::kable(tabla_coef,digits = 4)
```

## Count Effects in PHQ-9 Scores

```{r}
grafo_eff("Count")
```

## Zero effects in PHQ-9 Scores

```{r}
grafo_eff("Zero")
```

## What does that mean? Count part

- A unit increase in **BMI** increases the mean PHQ-9 score by a factor of 1%. 
- **Female** respondents have an increase of their PHQ-9 scores by a factor of 29%. 
- Each **stresor** event that happens to the respondant it's related with an increase their predicted PHQ-9 scores by a factor of 14%.
- **Age** and being an **Smoker** are not statistical significant for predicting phq-9 scores above 1.


## What does that mean? Zero part

- **Female** respondents have a decrease of chances of having zero symptoms by a factor 46%.
- Each **stressor** event in the last year that the respondent experiences decreases their chances of zero symptoms  by a factor of  40%.
- **Positive** life satisfaction increases the chances of having a zero PHQ-9 score by a factor of 2.36 more than baseline (Being **"Unsatisfied"**).
- **BMI**, **Neutral satisfaction** and **Smoker** are not relevant for predicting  a zero PHQ-9 score.

# Predicting PHQ-9 by ZINBR

## Predicting PHQ-9 Scores

Let's think about two people:

- **Person 1**:  45 years old, male. 0 stressors events in the last year. Satisfied with life. Non smoker.And with "Healthy Weight" (BMI=22).
- **Person 2**:  27 years old, female. 1 stressor event in the last year. Unsatisfied with life. Smoker. Moderate Obesity (BMI= 30).


```{r}
newobs1 <- tibble("s_imc"=22, 
                  "m0_sexo_fac"="Hombre",
                  "s13_estresores"=0,
                  "m0_edad"=45,
                  "s01_fac"="Satisfecho",
                  "s08_fac"= "No Fumador")

newobs2 <- tibble("s_imc"=30, 
                  "m0_sexo_fac"="Mujer",
                  "s13_estresores"=1,
                  "m0_edad"=27,
                  "s01_fac"="Insatisfecho",
                  "s08_fac"= "Fumador")

```

## Predicting PHQ-9 Scores by ZINB Regression

```{r}
tibble("x"=factor(0:27),
       "newobs1"=as.numeric(predict(mod_zinb, newdata = newobs1, type = "prob")),
       "newobs2"=as.numeric(predict(mod_zinb, newdata = newobs2, type = "prob")))%>%
  ggplot(aes(x=x,y=newobs1,color="#F8766D",fill="#F8766D",alpha=.3))+
  geom_col()+
  geom_col(aes(y=newobs2,color="#B79F00",fill="#B79F00",alpha=.3))+
  labs(fill=element_blank(),
       y="Probability",
       x="PHQ-9 Score")+
  scale_fill_discrete(labels=c("Person 2", "Person 1"))+
  scale_colour_discrete(guide="none")+
  scale_alpha(guide="none")+
  theme_classic()+
  theme(legend.position = "top")
```


## Future directions

- Assesing model fit for different probability distributions.
- Adding the longitudinal aspect of the model.
- Variable selection.
- Controlling missing data, adding sample weights, control panel attrition. 

## I'm happy to hear about

- Applications of PHQ-9 in longitudinal settings.
- Relevant factors in severity and persistance of depressive symptoms.

# Thank you!
